Identifying Effective Motivational Interviewing Sequences Using Automated Pattern Analysis

### 2018-04-13 jhir/latex-manuscript/manuscript.tex
update TABLES
fixed some spelling errors

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients’ motivational statements to guide clinical practice.

@@

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change~\cite{miller2013motivational}. The theory underlying MI’s clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}. The relational hypothesis suggests that counselors’ use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change. MI’s technical hypothesis~\cite{miller2009toward} states that counselors’ use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''. Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers’ ability to test MI’s technical hypothesis.

Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}. Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}. Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy. A parallel study of the adolescents’ caregivers~\cite{jacques2016building} drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk. Across these studies, counselors’ use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.

@@

HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}, while HMM was used to capture important characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient’s underlying motivational state, during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patient’s behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.

@@

In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI’s technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper provides the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al.~\cite{bertholet2010change} used HMM and adjusted regression models to examine the association of different sequences of ``change talk'' utterances with drinking outcomes in a brief motivational intervention. Their HMM model limited hidden states to three, which are ``towards change'', ``away from change'' and ``non-determined''. First and last hidden states decoded for a communication sequences, were used in conjunction with those three hidden states variables as predictors in a bivariate regression model, the results of which indicated that only the last state was significantly associated with drinking 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and Frequent Pattern Mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research. First-order Markov chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

@@

AF & Affirmation & Positive or complimentary statements that express appreciation, confidence, or reinforce the patient’s strengths or efforts. & ``You guys, as a family, are already doing a lot of really positive things.'' \\
AR & Action reflection & Statements that reflect back the patient’s statement(s) while at the same time embedding a solution to a barrier or an action plan. & ``If you decide to follow a meal plan, it has to include occasional dessert.'' \\
EA & Emphasize autonomy & Statements that directly acknowledge, honor, or emphasize the patient’s freedom of choice, autonomy, personal responsibility, and so forth. & ``Okay. Well, it’s your plan, so whatever works best for you. If you feel like you want one that’s written down that you can refer back to, then let’s write it and if not then that’s fine.''\\

@@

GINFOP & General information positive & The counselor gives advice, makes a suggestion, offers a solution/possible action, gives feedback, expresses a concern, or offers educational information in a patient-centered manner (i.e., asking permission, using the third person, giving the opportunity to reject the information, and offering a menu of options). & ``Okay.  Alright so I just wanted to tell you that I will be asking you a lot of questions.  It may get redundant.  So, if at any point in time you need a break or I’m asking too much go ahead and let me know.''\\

@@

QECHTP & Question to elicit change talk positive & Questions that ask about the patient’s desire, ability, reasons, or need for change or that reference past action toward behavior change or barriers to change. & ``Okay.  And tell me a little bit more about that.  Like what do you foresee your goal in this program?  Like what do you want to happen out of this program?''\\
QECMLP & Question to elicit commitment language positive & Questions that ask about current or future action toward behavior change or reference barriers to change. & ``Okay.  Is there something else that you could do eat maybe instead of a Pop-Tart that’s a little bit healthier?''\\
QEF & Question to elicit feedback & Statements that solicit the patient’s thoughts, ideas, or feelings about a specific recommendation or piece of information. & ``So, do you have any questions about that?''\\

@@

RCHTP & Reflect change talk positive & A reflective listening statement that captures and returns a patient’s statement or behavior from the current or a previous session that describes the patient’s desire, ability, reasons, or need for change or past action or barriers to change. & ``So, it sounds like you just want to be healthy and you want to be stylish. You want to fit into some different types of clothes.''\\
RCMLP & Reflect commitment language positive & A reflective listening statement that captures and returns a patient’s statement or behavior from the current or a previous session that describes current or future action or references barriers to changing with the goal of problem-solving. & ``You are ready to start this plan today.''\\
RO & Reflect other & A reflective listening statement that captures and returns a patient’s utterance or behavior from the current or previous session that is unrelated to the target behavior. & ``You are having a hard time at work.''\\

@@

SO & Statement other & An utterance eliciting feedback, offering support, self-disclosure, or of some other form besides a strategy or reflection & ``You’re being pulled in a million directions''\\

@@

SS & Structure session & A communication strategy that suggests an attempt to describe what will happen in the session or to refocus a meandering conversation back to the target behaviors & ``Maybe when the three of us come together in a few minutes, that’s something that we could just clarify with her, like is that really what she wants.''\\
SUM & Summary & A reflective listening statement that captures and returns at least 2 different ideas from a patient’s utterance or behavior from the current session & ``You have thought a lot about this.  Sometimes it feels like losing weight is just too hard.  Yet you have lots of reasons to lose weight.  If you could find a program you could stick to, a program that would not have too many changes at once, you would consider it.''\\

\rowcolor{lightgraycol}
\noalign{\smallskip}
\multicolumn{4}{l}{\textbf{Patient}} \\
\noalign{\smallskip}
CT & Change Talk & Statements that express the patient’s desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) changing their behavior & ``I will try to buy less junk food.''\\
ST & Sustain Talk & Statements that express the patient’s desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) to maintain the status quo or not change their behavior  & ``I didn’t get to the gym this week.''\\
HUPO	 & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
HUPW & High uptake weight & A turn that does develop the topic of the conversation. High Uptake statements include: weight related statements about actions of commitment, change talk, and ambivalence that occurred in the past, patient questions to the counselor, and session interruptions by persons who are not an active part of the treatment session. & ``Support is always good.  You know that’s a key factor.  Mm-hmm.''\\
LUP & Low uptake  & an utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
\hline
\end{longtable}
\end{small}

The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}. The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein’s conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\begin{itemize}
\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
\item $T$ is an state transition probability matrix, in which is the probability of HMM transitioning from state to state  
\item $E$ is an emission probability matrix, in which is the probability of observing symbol, when HMM is in state 
\item $p$ is the initial state distribution vector where is the probability of the first HMM state
\end{itemize}

@@

\begin{sidewaystable}
\vspace{32\baselineskip}

\caption{Hidden Markov Model Emission Matrices}
\label{tab:emission}
\begin{tabularx}{\textwidth}{lXXXp{0.4cm}XXXXXXXXXXXXXXXXXXX}
\hline\noalign{\smallskip}
State & AF & AR & EA & GIN- FON & GIN- FOP & QEB & QEC- HTP & QEC- MLP & QEF & QE- ST & QO & RC- HTP & RC- MLP & RO & RST & SO & SPT & SS & SUM & HU- PO & HU- PW & LUP \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\rowcolor{lightgraycol}
\multicolumn{23}{l}{\textbf{Successful  Sequences}} \\

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state’s emission probability mass, were used to interpret the emission matrix and label the hidden states. Most (84\%) successful sequences began with a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%). Successful sequences began with a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{Moderate receptivity} sequences were distinguished from \textit{low receptivity} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient’s autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

@@

\section{Conclusion}
\label{sec:conclusion}
These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods. While counselors’ use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data. The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.

>>>>

\begin{abstract}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change. Although there is strong empirical evidence linking ``MI-consistent'' counselor behaviors and patient motivational statements (i.e., ``change talk''), the specific counselor communication behaviors effective for eliciting patient change talk vary by treatment context and, thus, are a subject of ongoing research. A significant barrier to this research is the time and resource-intensive cognitive tasks, including the manual coding and sequential analysis of coded transcripts, that are an integral part of the qualitative analysis of patient-counselor communication. Data mining and machine learning techniques have the potential to significantly reduce this barrier by partially automating these cognitive tasks. In this paper, we evaluate the empirical effectiveness of the Hidden Markov Model, a probabilistic generative model for sequence data, for modeling sequences of behavior codes and closed frequent pattern mining, a method to identify frequently occurring sequential patterns of behavior codes in MI communication sequences to inform MI practice. We conducted experiments with 1,360 communication sequences from 37 transcribed audio recordings of weight loss counseling sessions with African-American adolescents with obesity and their caregivers. Transcripts had been previously annotated with patient-counselor behavior codes using a specialized codebook. Empirical results indicate that Hidden Markov Model and closed frequent pattern mining techniques can identify counselor communication strategies that are effective at eliciting patients' motivational statements to guide clinical practice.

@@

\section{Introduction}
\label{sec:intro}
Motivational Interviewing (MI) is an evidence-based strategy for communicating with patients about behavior change~\cite{miller2013motivational}. The theory underlying MI's clinical efficacy posits that behavior change is triggered by fostering an atmosphere of change, which is accomplished through the exercise of relational and technical skills~\cite{miller2013motivational}. The relational hypothesis suggests that counselors' use of accurate empathy, positive regard, and congruence create the ``spirit of MI'', an optimal therapeutic state to explore behavior change. MI's technical hypothesis~\cite{miller2009toward} states that counselors' use of communication techniques consistent with the MI framework (MICO; e.g., open-ended questions, reflections, advise with permission, affirmations, emphasize control, reframe, and support) will lead to patient ``change talk''. Change talk is patient statements during clinical encounters that express their internal desire, ability, reasons, need for, and/or commitment to behavior change. Previous studies~\cite{apodaca2009mechanisms} have shown that change talk expressed during treatment sessions consistently predicts behavior change with results persisting as long as 34 months post-intervention~\cite{walker2011influence}. In contrast, MI-inconsistent communication behaviors (MIIN; e.g., advising without permission, warning about behavioral consequences, and confronting) are hypothesized to lead to arguments against behavioral change and/or to maintain the status quo (referred to as counter change talk or sustain talk). Multiple studies have linked high rates of MICO to the expression of change talk and MIIN to sustain talk~\cite{magill2014technical}. These studies have relied on session-level behavior counts and correlational analyses, which ignore the temporal order of utterances in patient-counselor communication, thereby limiting researchers' ability to test MI's technical hypothesis.

Sequential analysis is an analytic approach to examine temporally ordered sequences of events or observations~\cite{bakeman1997observing, bakeman2011sequential}. Moyers and Martin~\cite{moyers2006therapist} applied sequential analysis in a study of adults in treatment for alcohol abuse and found change talk was significantly more likely after MICO and sustain talk more likely after MIIN. A follow up study with the same population found change talk was more likely after two MICO behaviors, counselor questions about the positive and negative aspects of drinking and reflections of change talk, but these behaviors also led to sustain talk~\cite{moyers2009session}. Surprisingly, MIIN was unrelated to sustain talk, but decreased the likelihood of change talk. Gaume et al.~\cite{gaume2008counsellor} used sequential analysis to study communication patterns during brief motivational interviewing for hazardous alcohol consumption with young adults conscripted into military service. They found MICO led to both change talk and sustain talk but the MIIN-to-sustain talk pattern was not observed. A second study with the same population confirmed that MICO leads to significantly more change talk and sustain talk~\cite{gaume2010counselor}. In this sample, MIIN led to greater sustain talk, but was unrelated to change talk. Further analyses revealed that reflections were the only MICO behavior linked to increased change talk; reflections and other MICO behaviors, excluding questions, were related to increased sustain talk. Glynn and colleagues~\cite{glynn2014change} linked reflections of change talk to the elicitation change talk and reflections of sustain talk to the elicitation of sustain talk among incarcerated adolescents with high rates of alcohol and marijuana use. In a study of adolescents engaged in weight loss treatment, Carcone et al.~\cite{carcone2013provider} used sequential analysis to identify three counselor behaviors likely to result in change talk: open-ended questions phrased to elicit change talk, reflections of change talk, and statements emphasizing decision-making autonomy. A parallel study of the adolescents' caregivers~\cite{jacques2016building} drew a similar conclusion, asking questions phrased to elicit change talk, reflections of change talk, and autonomy supportive statements were the counselor behaviors empirically linked to the elicitation of change talk. Across these studies, counselors' use of reflections was consistently linked to change talk; other MICO behaviors led to change talk in some contexts but not others suggesting a need for additional research to understand the contexts in which the various MICO strategies are effective.

@@

HMMs are widely used for the analysis of sequence data due to their ability to model short-range dependencies between clusters of discrete observations in a sequence. The HMM associates each observation in a sequence with a ``hidden'' state, which corresponds to a distribution over all distinct observations in a sequence (i.e., probabilities associated with each observation, when HMM is in this hidden state), such that each ``hidden'' corresponds to a different distribution. Sequences of observations are modeled as transitions between different hidden states and sampling observations from distributions corresponding to each hidden state. HMMs were originally proposed for speech recognition~\cite{rabiner1989tutorial}, in which the states were used to represent all English language sounds. In biomedical informatics, HMMs were employed for the diagnosis of diseases and biological sequence modeling~\cite{uuguz2007biomedical, amrhein2003client}. For example, Doppler ultrasound was used to extract features from imaging data and then an HMM-based classifier was applied to distinguish healthy patients from those with heart disease~\cite{uuguz2007biomedical}, while HMM was used to capture important characteristics of protein families~\cite{amrhein2003client}. In the application of HMM to patient-counselor communication, hidden states correspond to sets of related behavior codes, such as a patient's underlying motivational state, during a patient-counselor encounter. Although the MI literature has established patient change talk and commitment language (commitment language is a special class of change talk where patients express their intentions, plans, and action steps toward behavior change~\cite{baldi1994hidden}) as the antecedents of patient's behavior change~\cite{apodaca2009mechanisms}, there is less clarity regarding which counselor communication strategies influence the articulation of change talk. Modeling successful and unsuccessful communication sequences during MI sessions with HMM can provide additional evidence to identify the counselor communication strategies that likely to lead to patient change talk and commitment language.

@@

In this paper, we focus on computational methods to facilitate the sequential analysis of MI transcripts to identify patterns of patient-counselor communication in successful and unsuccessful sequences in MI sessions. Analysis of these patterns provides empirical support for the specific counselor communication strategies that are effective at eliciting patient change talk. This knowledge will inform MI theory by providing additional evidence to support MI's technical hypothesis. It will also inform clinical practice by facilitating the use of more effective and tailored counselor communication. This paper provides the first empirical evaluation of the effectiveness of closed frequent pattern mining to analyze patient-counselor communication sequences during MI sessions. Bertholet et al.~\cite{bertholet2010change} used HMM and adjusted regression models to examine the association of different sequences of ``change talk'' utterances with drinking outcomes in a brief motivational intervention. Their HMM model limited hidden states to three, which are ``towards change'', ``away from change'' and ``non-determined''. First and last hidden states decoded for a communication sequences, were used in conjunction with those three hidden states variables as predictors in a bivariate regression model, the results of which indicated that only the last state was significantly associated with drinking 12 months post-intervention. In the current study, we identified the optimal number of hidden states using HMM modeling of successful and unsuccessful sequences of patient-counselor communication. The goal of this study to evaluate the utility of using HMM and Frequent Pattern Mining to better understand the specific counselor communication strategies leading to patient change talk and sustain talk during Motivational Interviewing sessions. These two approaches offer the following advantages over the first-order Markov chain based methods most typically used in MI research. First-order Markov chain models identify the likely transitions between individual behaviors. In contrast, HMM summarizes transitions between clusters of related behavior codes (i.e., hidden states) allowing the identification of clusters of behaviors antecedent to change talk in successful patient-counselor communications and sustain talk in unsuccessful patient-counselor communications. Frequent pattern mining can identify patterns involving long-range dependencies between patient and counselor behaviors. Accounting for such long-range dependencies is important, since human behaviors, such as patient-counselor communications during MI sessions, are informed by all the antecedent behaviors and not just the immediately preceding behavior.

@@

AF & Affirmation & Positive or complimentary statements that express appreciation, confidence, or reinforce the patient's strengths or efforts. & ``You guys, as a family, are already doing a lot of really positive things.'' \\
AR & Action reflection & Statements that reflect back the patient's statement(s) while at the same time embedding a solution to a barrier or an action plan. & ``If you decide to follow a meal plan, it has to include occasional dessert.'' \\
EA & Emphasize autonomy & Statements that directly acknowledge, honor, or emphasize the patient's freedom of choice, autonomy, personal responsibility, and so forth. & ``Okay. Well, it's your plan, so whatever works best for you. If you feel like you want one that's written down that you can refer back to, then let's write it and if not then that's fine.''\\

@@

GINFOP & General information positive & The counselor gives advice, makes a suggestion, offers a solution/possible action, gives feedback, expresses a concern, or offers educational information in a patient-centered manner (i.e., asking permission, using the third person, giving the opportunity to reject the information, and offering a menu of options). & ``Okay.  Alright so I just wanted to tell you that I will be asking you a lot of questions.  It may get redundant.  So, if at any point in time you need a break or I'm asking too much go ahead and let me know.''\\

@@

QECHTP & Question to elicit change talk positive & Questions that ask about the patient's desire, ability, reasons, or need for change or that reference past action toward behavior change or barriers to change. & ``Okay.  And tell me a little bit more about that.  Like what do you foresee your goal in this program?  Like what do you want to happen out of this program?''\\
QECMLP & Question to elicit commitment language positive & Questions that ask about current or future action toward behavior change or reference barriers to change. & ``Okay.  Is there something else that you could do eat maybe instead of a Pop-Tart that's a little bit healthier?''\\
QEF & Question to elicit feedback & Statements that solicit the patient's thoughts, ideas, or feelings about a specific recommendation or piece of information. & ``So, do you have any questions about that?''\\

@@

RCHTP & Reflect change talk positive & A reflective listening statement that captures and returns a patient's statement or behavior from the current or a previous session that describes the patient's desire, ability, reasons, or need for change or past action or barriers to change. & ``So, it sounds like you just want to be healthy and you want to be stylish. You want to fit into some different types of clothes.''\\
RCMLP & Reflect commitment language positive & A reflective listening statement that captures and returns a patient's statement or behavior from the current or a previous session that describes current or future action or references barriers to changing with the goal of problem-solving. & ``You are ready to start this plan today.''\\
RO & Reflect other & A reflective listening statement that captures and returns a patient's utterance or behavior from the current or previous session that is unrelated to the target behavior. & ``You are having a hard time at work.''\\

@@

SO & Statement other & An utterance eliciting feedback, offering support, self-disclosure, or of some other form besides a strategy or reflection & ``You're being pulled in a million directions''\\

@@

SS & Structure session & A communication strategy that suggests an attempt to describe what will happen in the session or to refocus a meandering conversation back to the target behaviors & ``Maybe when the three of us come together in a few minutes, that's something that we could just clarify with her, like is that really what she wants.''\\
SUM & Summary & A reflective listening statement that captures and returns at least 2 different ideas from a patient's utterance or behavior from the current session & ``You have thought a lot about this.  Sometimes it feels like losing weight is just too hard.  Yet you have lots of reasons to lose weight.  If you could find a program you could stick to, a program that would not have too many changes at once, you would consider it.''\\

\rowcolor{lightgraycol}
\noalign{\smallskip}
\multicolumn{4}{l}{\textbf{Patient}} \\
\noalign{\smallskip}
CT & Change Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) changing their behavior & ``I will try to buy less junk food.''\\
ST & Sustain Talk & Statements that express the patient's desire, ability, reasons, need for, or commitment to (intentions, plans, and action steps) to maintain the status quo or not change their behavior  & ``I didn't get to the gym this week.''\\
HUPO	 & High uptake other & An utterance that develops the topic of the conversation but is about non-target behaviors or interruptions & ``Yeah because the mentor comes and they take off and they go someplace for a little while.''\\
HUPW & High uptake weight & A turn that does develop the topic of the conversation. High Uptake statements include: weight related statements about actions of commitment, change talk, and ambivalence that occurred in the past, patient questions to the counselor, and session interruptions by persons who are not an active part of the treatment session. & ``Support is always good.  You know that's a key factor.  Mm-hmm.''\\
LUP & Low uptake  & an utterance that does not develop the topic of conversation but still allows it to continue & ``Mm-hmm.  Right.''\\
\hline
\end{longtable}
\end{small}

The data set for this study consists of 37 audio recorded transcripts from Motivational Interviewing (MI) sessions conducted with African American adolescents seeking weight loss. Each transcript represented a one-on-one interaction between the adolescent and the MI counselor followed by a one-on-one interaction between the caregiver and the MI counselor. These encounters had previously been annotated using the Minority Youth Sequential Code for Observing Process Exchanges (MYSCOPE)~\cite{carcone2013provider}. The MYSCOPE is an adaptation of the original MI-SCOPE~\cite{martin2005motivational}, a qualitative code scheme to characterize patient-counselor communication during MI treatment sessions. The MYSCOPE was informed by MI fidelity code schemes including the MI Treatment Integrity Scale (MITI)~\cite{moyers2005assessing}, the MI Skill Code (MISC)~\cite{apodaca2007motivational}, and Amrhein's conceptualization of change talk and commitment~\cite{amrhein2004does}. A primary coder used the MYSCOPE to annotate transcripts of all 37 sessions; a secondary coder coded 20\% ($n = 7$) of the sessions for inter-rater reliability which was good ($k = .696$). The experimental dataset consists of 7,192 patient, caregiver, and counselor utterances segmented and annotated with the MYSCOPE behavior codes, illustrated in Table~\ref{tab:codebook}.

@@

\begin{itemize}
\item $M$ is the number of distinct observations symbols per state, i.e. the discrete codebook size (Table~\ref{tab:codebook})
\item $T$ is a state transition probability matrix, in which is the probability of HMM transitioning from state to state  
\item $E$ is an emission probability matrix, in which is the probability of observing symbol, when HMM is in state 
\item $p$ is the initial state distribution vector where is the probability of the first HMM state
\end{itemize}

@@

\begin{sidewaystable}
\vspace{43\baselineskip}

\caption{Hidden Markov Model Emission Matrices}
\label{tab:emission}
\begin{tabularx}{\textwidth}{lXXXp{0.4cm}XXXXXXXXXXXXXXXXXXX}
\hline\noalign{\smallskip}
State & AF & AR & EA & GIN-FON & GIN-FOP & QEB & QEC-HTP & QEC-MLP & QEF & QE-ST & QO & RC-HTP & RC-MLP & RO & RST & SO & SPT & SS & SUM & HU-PO & HU-PW & LUP \\
\noalign{\smallskip}\hline\noalign{\smallskip}
\rowcolor{lightgraycol}
\multicolumn{23}{l}{\textbf{Successful  Sequences}} \\

@@

\section{Results}
\label{sec:results}
The transition and emission probability matrices of HMM model are reported in Tables~\ref{tab:emission} and \ref{tab:transition}. The three behaviors with the highest frequency, representing 45-69\% of each state's emission probability mass, were used to interpret the emission matrix and label the hidden states. Most (84\%) successful sequences began with a state characterized as \textit{``high motivation''} as evidenced by a greater proportion of three counselor behaviors: reflections of change talk (29\%), reflections of commitment language (17\%), and affirmations (12\%). Successful sequences began with a state of \textit{``high receptivity''} 11\% of the time. \textit{``High receptivity''} sequences were characterized by nearly equal proportions of information offered using patient-centered strategies (18\%), questions to elicit change talk (16\%), and affirmations (15\%). Few successful sequences began in states of \textit{``moderate receptivity''} and \textit{``low receptivity''} (2\% and 3\% of the time, respectively). These two states were characterized by different proportions of the same behaviors. \textit{Moderate receptivity} sequences were distinguished from \textit{low receptivity} sequences by a greater proportion of counselor questions to elicit change talk (20\% versus 11\%) and a lower proportion of patient low uptake statements (16\% versus 28\%); counselor statements emphasizing the patient's autonomy were about the same (14\% versus 17\%). No (0\%) successful sequence began in the \textit{``active feedback''} state, which was characterized by three patient behaviors, low uptake (47\%), weight-related high uptake (12\%), and other-related high uptake (10\%). Successful sequences transitioned from \textit{``high motivation''} to \textit{``active feedback''} most often (41\%). \textit{``active feedback''}, in turn, most frequently transitioned to \textit{``high receptivity''} (39\%). The \textit{``moderate receptivity''} state transitioned to most often to \textit{``active feedback''} (27\%) and automatically transitioned back to \textit{``moderate receptivity''} (24\%) or to \textit{``low receptivity''} (23\%) with similar frequency. The full transition matrix is presented in Table~\ref{tab:transition}.

@@

\section{Conclusion}
\label{sec:conclusion}
These results add to the growing evidence base examining the mechanisms of effect in Motivational Interviewing using modeling approaches that overcome critical shortcomings of previous methods. While counselors' use of ``MI-consistent'' communication behaviors has been previously linked to higher rates of change talk in correlational studies~\cite{moyers2006therapist, catley2006adherence, thrasher2006motivational, mccambridge2011fidelity} and simple Markov chain models~\cite{moyers2006therapist, moyers2009session, gaume2010counselor}, the use of HMM and frequent pattern mining analyses improves upon these approaches by considering long-term dependencies in the data. The results of this work suggest a more complex pattern between counselor communication behaviors and patient talk that varies depending on the context in which Motivational Interviewing is being used.
